---
title: "SURV 727 Project"
author: "Talia Kaatz and Chris Brehm"
date: "12/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(knitr)
library(DBI)
library(dbplyr)
library(bigrquery)
library(sf)
library(magrittr)

```

## Initializing project in github and rmarkdown file

read in data from git repository

```{r, import data}

county_hv <- read_csv("https://raw.githubusercontent.com/taliakaatz/surv727project/master/county%20HV.csv")
county_rv <- read_csv("https://raw.githubusercontent.com/taliakaatz/surv727project/master/county%20RV.csv")
nhood_hv <- read_csv("https://raw.githubusercontent.com/taliakaatz/surv727project/master/neighborhood%20HV.csv")
nhood_rv <- read_csv("https://raw.githubusercontent.com/taliakaatz/surv727project/master/neighborhood%20RV.csv")


```


subset data to metro area (county level) and baltimore city (neighborhood level) and reshape to long format with full date, year, and month variables. also clean some superfluous variables



```{r, preprocess zillow data}

RVmetro <- 
  county_rv %>% 
    filter(Metro %in% "Baltimore-Columbia-Towson") %>%
    gather(key = "date", value = "RentalValue", 8:117) %>%
    select(- State, -Metro, -StateCodeFIPS) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    arrange(RegionName, desc(date))

HVmetro <- 
  county_hv %>% 
    filter(Metro %in% "Baltimore-Columbia-Towson") %>%
    gather(key = "date", value = "HomeValue", 8:290) %>%
    select(- State, -Metro, -StateCodeFIPS) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    arrange(RegionName, desc(date))

RVbmore  <-
  nhood_rv %>% 
    filter(City %in% "Baltimore" & State %in% "MD") %>%
    gather(key = "date", value = "RentalValue", 8:117) %>%
    select(-State, -Metro, -City, -CountyName) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    arrange(RegionName, desc(date))

HVbmore <-
  nhood_hv %>% 
    filter(City %in% "Baltimore") %>%
    gather(key = "date", value = "HomeValue", 8:290) %>%
    select(-State, -Metro, -City, -CountyName) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    arrange(RegionName, desc(date))

```


begin exploring data visualizations -- work in progress

```{r}

#line plot of home values in baltimore city over the course of 2018, broken out by neighborhood
HVbmore %>%
  filter(year == "2018") %>%
  ggplot() +
  geom_line(aes(x = month, y = HomeValue, group = RegionName, color = HomeValue))



```


first attempts at mapping visualizations -- work in progress


```{r}

#pull MD county map data from 'maps' package
md_map <- map_data("county", region = "maryland")

#map filling MD counties by average rental values in Jan 2019
county_rv %>% 
    filter(State %in% "MD") %>%
    gather(key = "date", value = "RentalValue", 8:117) %>%
    select(- State, -StateCodeFIPS) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    mutate(subregion = gsub(" County", "", RegionName)) %>%
    mutate(subregion = tolower(subregion)) %>%
    right_join(md_map, by = "subregion") %>%
    filter(date == "2019-01") %>%
    ggplot() + 
      geom_polygon(aes(x = long, y = lat, group = group, fill = RentalValue))

#map filling MD counties by average home values in Jan 2019
county_hv %>% 
    filter(State %in% "MD") %>%
    gather(key = "date", value = "HomeValue", 8:290) %>%
    select(- State, -StateCodeFIPS) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    mutate(subregion = gsub(" County", "", RegionName)) %>%
    mutate(subregion = tolower(subregion)) %>%
    right_join(md_map, by = "subregion") %>%
    filter(date == "2019-01") %>%
    ggplot() + 
      geom_polygon(aes(x = long, y = lat, group = group, fill = HomeValue))


#read in shapefile as a "simple feature"    
bmore_sf <- st_read("/Users/taliakaatz/Desktop/Computation & Data Display/Project Folder/baltimore city shape files/Maryland_Baltimore_City_Neighborhoods.shp")

#plot each aspect of shapefile
plot(bmore_sf)

##attempted to extract long/lat coordinates as X,Y values to append to zillow data and use in ggplot, was not successful
#sf1 <- bmore_sf[1,]
#sf1 <- st_cast(sf1, "POINT")
#st_write(sf1, "/Users/taliakaatz/Desktop/Computation & Data Display/Project Folder/baltimore city shape files/geo baltimore city.csv", layer_options = "GEOMETRY=AS_XY")
  
#bmore_map <- read.csv("/Users/taliakaatz/Desktop/Computation & Data Display/Project Folder/baltimore city shape files/geo baltimore city.csv")

#clean neighborhood names 

RVbmore %<>% 
  mutate(NHOOD = trimws(toupper(RegionName)))

bmore_sf %<>%
  mutate(NHOOD = trimws(toupper(LABEL)))

#check the shapefile names against the zillow names

hoodnames <- unique(RVbmore$NHOOD)
hoodnames[!(hoodnames %in% bmore_sf$NHOOD)]

othhoodnames <- unique(bmore_sf$NHOOD)
othhoodnames[!(othhoodnames %in% RVbmore$NHOOD)]

#adjust non-matches (zillow neighborhood names not matched to any in shaefile) as necessary and check results

RVbmore %<>%
  mutate(NHOOD = gsub("COPPIN HEIGHTS - ASH-CO-EAST", "COPPIN HEIGHTS/ASH-CO-EAST", NHOOD)) %>%
  mutate(NHOOD = gsub("ELLWOOD PARK-MONUMENT", "ELLWOOD PARK/MONUMENT", NHOOD)) %>%
  mutate(NHOOD = gsub("GLENHAM-BELFORD", "GLENHAM-BELHAR", NHOOD)) %>%
  mutate(NHOOD = gsub("MONDAWIN", "MONDAWMIN", NHOOD)) %>%
  mutate(NHOOD = gsub("NEW SOUTHWEST - MOUNT CLARE", "NEW SOUTHWEST/MOUNT CLARE", NHOOD)) %>%
  mutate(NHOOD = gsub("PENROSE-FAYETTE STREET OUTREACH", "PENROSE/FAYETTE STREET OUTREACH", NHOOD)) %>%
  mutate(NHOOD = gsub("SBIC", "SOUTH BALTIMORE", NHOOD)) %>%
  mutate(NHOOD = gsub("WASHINGTON VILLAGE", "WASHINGTON VILLAGE/PIGTOWN", NHOOD)) %>%
  mutate(NHOOD = gsub("WOODRING", "WESTFIELD", NHOOD))

hoodnames <- unique(RVbmore$NHOOD)
hoodnames[!(hoodnames %in% bmore_sf$NHOOD)]
##based on output, still need to resolve "HARFORD-ECHODALE - PERRING PARKWAY" -- need to determine which neighborhood that area falls under in shapefile

##examples of how to plot a single attribute of shapefile using 'plot' and 'ggplot', other output specifications (fill, etc.) can be added -- neither of these display useful info yet, they just color-code each neighborhood individually by the neighborhood name

#using plot
plot(bmore_sf['LABEL'])

#using ggplot
bmore_sf %>%
  ggplot() +
    geom_sf(aes(fill = LABEL), show.legend = FALSE)
  


```

