---
title: "SURV 727 Final Report"
author: "Talia Kaatz and Chris Brehm"
date: "12/13/2019"
output: html_document
---
#Introduction

Housing affordability and changes in the housing market garner significant attention from politicians, activists, and economists, and can be seen as indicators of broader social issues. Gentrification, predatory landlords and lenders, and segregation of housing markets as indicators of racial inequality and oppression, societal fears of economic instability manifest in predictions of the next bubble burst or housing shortages, and generational changes can be indicated by shifts in renting vs. ownership.

Of the many American metropolitan areas affected by the housing crisis, the city of Baltimore, MD provides distinct characteristics that uniquely qualify it to study as a case study. Baltimore City, unlike in many major metro areas, is not part of any surrounding counties but is classed as an “independent city.” Tax bases are therefore separate at both city and county level, providing unique comparison of urban/suburban areas, as well as providing a basis for comparison between the city and other Maryland counties. Moreover, Baltimore’s history of redlining and segregation within the city has produced stark demographic and economic differences at the neighborhood level, and a recent study ranked Baltimore among most heavily-gentrified cities in U.S. (Meehan 2019).

In order to better understand the housing crisis and the factors associated with it, we have drawn data from a variety of sources including Zillow, the American Community Survey, and the Baltimore Neighborhood Indicator Alliance. Using this data, we will investigate the following hypotheses:

H1: County-level rental values will show a greater increase over time than housing values over the same period

H2: This effect will be more pronounced in urban than in rural counties

H3: Areas with younger populations will have higher rental values (compared to housing values) than areas with older populations

H4: Increases over time in home and rental values will be associated with decreases in the proportion of non-white residents

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(knitr)
library(DBI)
library(dbplyr)
library(bigrquery)
library(sf)
library(magrittr)
library(censusapi)
library(readxl)

```

#Background

The most recent edition of the Oxford English Dictionary defines gentrification as “the process of renovating and improving housing or a district so that it conforms to middle-class taste” (Lexico 2019). Advocates and city residents often express concerns about gentrification, as it is typically associated with the displacement of poorer or non-white residents from these areas (Capps 2019, Meehan 2019). Additionally, rising rents and lack of home-ownership among millenials have led to growing concerns about rental markets, in addition to home sales markets (Hoffower 2019, Kasakove 2019). 

However, gentrification is not always (perhaps not even usually) strongly associated with displacement or negative effects for minority residents (Brummet and Reed 2019). Understanding housing equity is of particular importance in the U.S., where racial segregation and housing policy have been used to restrict minority groups’ access to wealth and resources (Squire and Cubrin 2005, Hoynes and McFadden 1994). Due to the complex relationships at play between time, geography, and various community demographics and characteristics, this is an area where creative data visualizations can aid in exploring and understanding patterns of housing costs, race, and age. Comparing race and age distributions in an area with average housing values or rents may each reveal a different picture of these relationships. 

#Data



```{r, import data}

#Zillow data first, from the github repository
county_hv <- read_csv("https://raw.githubusercontent.com/taliakaatz/surv727project/master/county%20HV.csv")
county_rv <- read_csv("https://raw.githubusercontent.com/taliakaatz/surv727project/master/county%20RV.csv")
nhood_hv <- read_csv("https://raw.githubusercontent.com/taliakaatz/surv727project/master/neighborhood%20HV.csv")
nhood_rv <- read_csv("https://raw.githubusercontent.com/taliakaatz/surv727project/master/neighborhood%20RV.csv")

#ACS data next, using 'censusapi'
cs_key <- "a57bdfaf40991f97bfd12e41478426c81badd18a"
count_md18 <- getCensus(name = "2018/acs/acs1/profile",
                      vars = c("DP05_0037PE", "DP05_0071PE", "DP05_0018E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]

count_md17 <- getCensus(name = "2017/acs/acs1/profile",
                      vars = c("DP05_0037PE", "DP05_0071PE", "DP05_0018E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]

count_md16 <- getCensus(name = "2016/acs/acs1/profile",
                      vars = c("DP05_0032PE", "DP05_0066PE", "DP05_0017E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]

count_md15 <- getCensus(name = "2015/acs/acs1/profile",
                      vars = c("DP05_0032PE", "DP05_0066PE", "DP05_0017E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]

count_md14 <- getCensus(name = "2014/acs/acs1/profile",
                      vars = c("DP05_0032PE", "DP05_0066PE", "DP05_0017E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]

count_md13 <- getCensus(name = "2013/acs/acs1/profile",
                      vars = c("DP05_0032PE", "DP05_0066PE", "DP05_0017E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]

count_md12 <- getCensus(name = "2012/acs/acs1/profile",
                      vars = c("DP05_0032PE", "DP05_0066PE", "DP05_0017E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]

count_md11 <- getCensus(name = "2011/acs/acs1/profile",
                      vars = c("DP05_0032PE", "DP05_0066PE", "DP05_0017E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]

count_md10 <- getCensus(name = "2010/acs/acs1/profile",
                      vars =c("DP05_0032PE", "DP05_0066PE", "DP05_0017E"),#%white, %hisp, median age
                      region = "county:*", 
                      regionin = "state:24",
                      key = cs_key)[-c(1:2)]


#BNIA data, also from the guthub repository
bnia_crosswalk <- read.csv("/Users/taliakaatz/Desktop/Computation & Data Display/Project Folder/CSA_to_NSA_2010.csv")

#demographic data, read in as a separate dataset for each year
bnia <- "/Users/taliakaatz/Desktop/Computation & Data Display/Project Folder/VS17_Indicators_2010_2017.xlsx"

bnia2010 <- read_xlsx(bnia, sheet = "2010")
bnia2011 <- read_xlsx(bnia, sheet = "2011")
bnia2012 <- read_xlsx(bnia, sheet = "2012")
bnia2013 <- read_xlsx(bnia, sheet = "2013")
bnia2014 <- read_xlsx(bnia, sheet = "2014")
bnia2015 <- read_xlsx(bnia, sheet = "2015")
bnia2016 <- read_xlsx(bnia, sheet = "2016")
bnia2017 <- read_xlsx(bnia, sheet = "2017")

#shapefile for mapping
bnia_sf <- st_read("/Users/taliakaatz/Desktop/Computation & Data Display/Project Folder/csa_2010_boundaries/CSA_NSA_Tracts.shp")

#MD county map data, from 'maps' package
md_map <- map_data("county", region = "maryland")


```

#Preprocessing
##County-level data
```{r, prepare county-level Zillow files}

RVMD <-
  county_rv %>% 
    filter(State %in% "MD") %>%
    gather(key = "date", value = "RentalValue", 8:117) %>%
    select(-State, -StateCodeFIPS, -MunicipalCodeFIPS, -SizeRank) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    filter(year %in% 2010:2018) %>%
    arrange(RegionName, desc(date))

HVMD <-
  county_hv %>% 
    filter(State %in% "MD") %>%
    gather(key = "date", value = "HomeValue", 8:290) %>%
    select(-State, -StateCodeFIPS, -MunicipalCodeFIPS, -SizeRank) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    filter(year %in% 2010:2018) %>%
    arrange(RegionName)
```

```{r, prepare ACS county data}


count_md10$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")

count_md11$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")

count_md12$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")

count_md13$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")

count_md14$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")

count_md15$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")

count_md16$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")

count_md17$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")

count_md18$CountyName <- c("Allegany County","Anne Arundel County", "Baltimore County", "Calvert County", "Carroll County", "Cecil County", "Charles County", "Frederick County", "Harford County", "Howard County", "Montgomery County", "Prince George's County", "Saint Mary's County", "Washington County", "Wicomico County", "Baltimore City")



```

```{r, write ACS data to new variables in Zillow data files}
HVMD$nonwhite <- NA
HVMD$hispanic <- NA
HVMD$medianage <- NA
RVMD$nonwhite <- NA
RVMD$hispanic <- NA
RVMD$medianage <- NA

for(i in 1:length(HVMD$RegionName)){
  if(HVMD$RegionName[i] %in% count_md10$CountyName){
    if(HVMD$year[i] == 2010){
      HVMD$nonwhite[i] <- 100 - count_md10$DP05_0032PE[count_md10$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md10$DP05_0066PE[count_md10$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md10$DP05_0017E[count_md10$CountyName == HVMD$RegionName[i]]
    }
    else if(HVMD$year[i] == 2011){
      HVMD$nonwhite[i] <- 100 - count_md11$DP05_0032PE[count_md11$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md11$DP05_0066PE[count_md11$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md11$DP05_0017E[count_md11$CountyName == HVMD$RegionName[i]]
    }
    else if(HVMD$year[i] == 2012){
      HVMD$nonwhite[i] <- 100 - count_md12$DP05_0032PE[count_md12$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md12$DP05_0066PE[count_md12$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md12$DP05_0017E[count_md12$CountyName == HVMD$RegionName[i]]
    }
    else if(HVMD$year[i] == 2013){
      HVMD$nonwhite[i] <- 100 - count_md13$DP05_0032PE[count_md13$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md13$DP05_0066PE[count_md13$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md13$DP05_0017E[count_md13$CountyName == HVMD$RegionName[i]]
    }
    else if(HVMD$year[i] == 2014){
      HVMD$nonwhite[i] <- 100 - count_md14$DP05_0032PE[count_md14$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md14$DP05_0066PE[count_md14$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md14$DP05_0017E[count_md14$CountyName == HVMD$RegionName[i]]
    }
    else if(HVMD$year[i] == 2015){
      HVMD$nonwhite[i] <- 100 - count_md15$DP05_0032PE[count_md15$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md15$DP05_0066PE[count_md15$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md15$DP05_0017E[count_md15$CountyName == HVMD$RegionName[i]]
    }
    else if(HVMD$year[i] == 2016){
      HVMD$nonwhite[i] <- 100 - count_md16$DP05_0032PE[count_md16$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md16$DP05_0066PE[count_md16$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md16$DP05_0017E[count_md16$CountyName == HVMD$RegionName[i]]
    }
    else if(HVMD$year[i] == 2017){
      HVMD$nonwhite[i] <- 100 - count_md17$DP05_0037PE [count_md17$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md17$DP05_0071PE[count_md17$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md17$DP05_0018E[count_md17$CountyName == HVMD$RegionName[i]]
    }
    else if(HVMD$year[i] == 2018){
      HVMD$nonwhite[i] <- 100 - count_md18$DP05_0037PE[count_md18$CountyName == HVMD$RegionName[i]]
      HVMD$hispanic[i] <- count_md18$DP05_0071PE[count_md18$CountyName == HVMD$RegionName[i]]
      HVMD$medianage[i] <- count_md18$DP05_0018E[count_md18$CountyName == HVMD$RegionName[i]]
    }
    
  }
}

for(i in 1:length(RVMD$RegionName)){
  if(RVMD$RegionName[i] %in% count_md10$CountyName){
    if(RVMD$year[i] == 2010){
      RVMD$nonwhite[i] <- 100 - count_md10$DP05_0032PE[count_md10$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md10$DP05_0066PE[count_md10$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md10$DP05_0017E[count_md10$CountyName == RVMD$RegionName[i]]
    }
    else if(RVMD$year[i] == 2011){
      RVMD$nonwhite[i] <- 100 - count_md11$DP05_0032PE[count_md11$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md11$DP05_0066PE[count_md11$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md11$DP05_0017E[count_md11$CountyName == RVMD$RegionName[i]]
    }
    else if(RVMD$year[i] == 2012){
      RVMD$nonwhite[i] <- 100 - count_md12$DP05_0032PE[count_md12$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md12$DP05_0066PE[count_md12$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md12$DP05_0017E[count_md12$CountyName == RVMD$RegionName[i]]
    }
    else if(RVMD$year[i] == 2013){
      RVMD$nonwhite[i] <- 100 - count_md13$DP05_0032PE[count_md13$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md13$DP05_0066PE[count_md13$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md13$DP05_0017E[count_md13$CountyName == RVMD$RegionName[i]]
    }
    else if(RVMD$year[i] == 2014){
      RVMD$nonwhite[i] <- 100 - count_md14$DP05_0032PE[count_md14$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md14$DP05_0066PE[count_md14$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md14$DP05_0017E[count_md14$CountyName == RVMD$RegionName[i]]
    }
    else if(RVMD$year[i] == 2015){
      RVMD$nonwhite[i] <- 100 - count_md15$DP05_0032PE[count_md15$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md15$DP05_0066PE[count_md15$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md15$DP05_0017E[count_md15$CountyName == RVMD$RegionName[i]]
    }
    else if(RVMD$year[i] == 2016){
      RVMD$nonwhite[i] <- 100 - count_md16$DP05_0032PE[count_md16$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md16$DP05_0066PE[count_md16$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md16$DP05_0017E[count_md16$CountyName == RVMD$RegionName[i]]
    }
    else if(RVMD$year[i] == 2017){
      RVMD$nonwhite[i] <- 100 - count_md17$DP05_0037PE [count_md17$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md17$DP05_0071PE[count_md17$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md17$DP05_0018E[count_md17$CountyName == RVMD$RegionName[i]]
    }
    else if(RVMD$year[i] == 2018){
      RVMD$nonwhite[i] <- 100 - count_md18$DP05_0037PE[count_md18$CountyName == RVMD$RegionName[i]]
      RVMD$hispanic[i] <- count_md18$DP05_0071PE[count_md18$CountyName == RVMD$RegionName[i]]
      RVMD$medianage[i] <- count_md18$DP05_0018E[count_md18$CountyName == RVMD$RegionName[i]]
    }
    
  }
}
HVMD$nonwhite[HVMD$nonwhite > 100] <- NA
RVMD$nonwhite[RVMD$nonwhite > 100] <- NA

combinedMD <- 
  HVMD %>%
    left_join(RVMD, by = c("RegionName", "date"))

```
#Preprocessing
##Neighborhood-level data 
```{r, prepare neighborhood-level Zillow data}

RVBM  <-
  nhood_rv %>% 
    filter(City %in% "Baltimore" & State %in% "MD") %>%
    gather(key = "date", value = "RentalValue", 8:117) %>%
    select(-State, -Metro, -City, -CountyName) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    filter(year %in% 2010:2018) %>%
    arrange(RegionName, desc(date))

HVBM <-
  nhood_hv %>% 
    filter(City %in% "Baltimore") %>%
    gather(key = "date", value = "HomeValue", 8:290) %>%
    select(-State, -Metro, -City, -CountyName) %>%
    separate(date, into = c("year", "month"), sep = "-", remove = FALSE) %>%
    filter(year %in% 2010:2018) %>%
    arrange(RegionName, desc(date))

#check whether neighborhood names line up between rental and home value files

countynames <- unique(RVBM$RegionName)
rent_hoods <- countynames[!(countynames %in% HVBM$RegionName)]

countynames2 <- unique(HVBM$RegionName)
home_hoods <- countynames2[!(countynames2 %in% RVBM$RegionName)]

print(rent_hoods)
print(home_hoods)
#evidence of significant differences in the neighborhoods present in each file

```

Standardize crosswalk and zillow files and merge by neighborhood

```{r, merge neighborhood crosswalk file to Zillow data}

#create uniform neighborhood name variables in each dataset
  
bnia_crosswalk %<>% 
  mutate(NHOOD = trimws(toupper(NSA2010))) %>%
  select(-NSA2010)
  
HVBM %<>%
  mutate(NHOOD = trimws(toupper(RegionName)))

RVBM %<>%
  mutate(NHOOD = trimws(toupper(RegionName))) 

#merge crosswalk to home and rental value data and identify unmatched neighborhood names
HVBM %>%
  left_join(bnia_crosswalk, by = "NHOOD") %>%
  filter(is.na(CSA2010)) %>%
  distinct(NHOOD)

RVBM %>%
  left_join(bnia_crosswalk, by = "NHOOD") %>%
  filter(is.na(CSA2010)) %>%
  distinct(NHOOD)

#fix unmatched neighborhood names (update in Zillow file to match crosswalk): 

HVBM %<>%
  mutate(NHOOD = gsub("CHINQUAPIN PARK-BELVEDERE", "CHINQUAPIN PARK", NHOOD)) %>%
  mutate(NHOOD = gsub("HARFORD-ECHODALE - PERRING PARKWAY", "HARFORD-ECHODALE/PERRING PARKWAY", NHOOD)) %>%
  mutate(NHOOD = gsub("JOSEPH LEE", "BAYVIEW", NHOOD)) %>%
  mutate(NHOOD = gsub("MOUNT WASHINGTON", "MT. WASHINGTON", NHOOD)) %>%
  mutate(NHOOD = gsub("NORTH ROLAND PARK - POPLAR HILL", "NORTH ROLAND PARK/POPLAR HILL", NHOOD)) %>%
  mutate(NHOOD = gsub("SBIC", "SOUTH BALTIMORE", NHOOD)) %>%
  mutate(NHOOD = gsub("WOODRING", "WESTFIELD", NHOOD))


RVBM %<>%
  mutate(NHOOD = gsub("COPPIN HEIGHTS - ASH-CO-EAST", "COPPIN HEIGHTS/ASH-CO-EAST", NHOOD)) %>%
  mutate(NHOOD = gsub("ELLWOOD PARK-MONUMENT", "ELLWOOD PARK/MONUMENT", NHOOD)) %>%
  mutate(NHOOD = gsub("HARFORD-ECHODALE - PERRING PARKWAY", "HARFORD-ECHODALE/PERRING PARKWAY", NHOOD)) %>%
  mutate(NHOOD = gsub("MONDAWIN", "MONDAWMIN", NHOOD)) %>%
  mutate(NHOOD = gsub("MOUNT WASHINGTON", "MT. WASHINGTON", NHOOD)) %>%
  mutate(NHOOD = gsub("NEW SOUTHWEST - MOUNT CLARE", "NEW SOUTHWEST/MT. CLARE", NHOOD)) %>%
  mutate(NHOOD = gsub("PENROSE-FAYETTE STREET OUTREACH", "PENROSE/FAYETTE STREET OUTREACH", NHOOD)) %>%
  mutate(NHOOD = gsub("SBIC", "SOUTH BALTIMORE", NHOOD)) %>%
  mutate(NHOOD = gsub("WASHINGTON VILLAGE", "WASHINGTON VILLAGE/PIGTOWN", NHOOD)) %>%
  mutate(NHOOD = gsub("WOODRING", "WESTFIELD", NHOOD))

#merge and check again with updated neighborhoods -- looks good! 

HVBM %>%
  left_join(bnia_crosswalk, by = "NHOOD") %>%
  filter(is.na(CSA2010)) %>%
  distinct(NHOOD)

RVBM %>%
  left_join(bnia_crosswalk, by = "NHOOD") %>%
  filter(is.na(CSA2010)) %>%
  distinct(NHOOD)

#merge CSA names from crosswalk to datasets

HVBM %<>% left_join(bnia_crosswalk, by = "NHOOD")

RVBM %<>% left_join(bnia_crosswalk, by = "NHOOD")

#adjust some CSA names to match formatting to BNIA demographic files

RVBM %<>%
  mutate(CSA2010 = gsub("Allendale/Irvington/South Hilton", "Allendale/Irvington/S. Hilton", CSA2010)) %>%
  mutate(CSA2010 = gsub("Glen-Falstaff", "Glen-Fallstaff", CSA2010)) %>%
  mutate(CSA2010 = gsub("Westport/Mt. Winans/Lakeland", "Westport/Mount Winans/Lakeland", CSA2010)) %>%
  mutate(CSA2010 = gsub("Mt. Washington/Coldspring", "Mount Washington/Coldspring", CSA2010))

HVBM %<>%
  mutate(CSA2010 = gsub("Allendale/Irvington/South Hilton", "Allendale/Irvington/S. Hilton", CSA2010)) %>%
  mutate(CSA2010 = gsub("Glen-Falstaff", "Glen-Fallstaff", CSA2010)) %>%
  mutate(CSA2010 = gsub("Westport/Mt. Winans/Lakeland", "Westport/Mount Winans/Lakeland", CSA2010)) %>%
  mutate(CSA2010 = gsub("Mt. Washington/Coldspring", "Mount Washington/Coldspring", CSA2010))

```

Prepare BNIA demographic files for merge

```{r, prepare BNIA demographic files}

#in file for each year, subset to variables of interest and rename

#2010
bnia_vars <- colnames(bnia2010[, c("CSA2010", "Percent of Residents - White/Caucasian (Non-Hispanic)", "Percent of Residents - Hispanic", "Percent of Population 18-24 Years old", "Median Price of Homes Sold", "Affordability Index - Mortgage", "Affordability Index - Rent")])
new_varnames <- c("CSA2010", "nonwhite10", "hispanic10", "percentya10", "mediansold10", "ai_mortgage10", "ai_rent10")

bnia2010 %<>%
  select(bnia_vars) %>%
  set_colnames(new_varnames) %>%
  mutate(nonwhite10 = (100 - nonwhite10))

#2011
bnia_vars <- c("CSA2010", "Median Price of Homes Sold", "Affordability Index - Mortgage", "Affordability Index - Rent")
new_varnames <- c("CSA2010", "mediansold11", "ai_mortgage11", "ai_rent11")

bnia2011 %<>%
  select(bnia_vars) %>%
  set_colnames(new_varnames)

#2012
bnia_vars <- c("CSA2010", "Median Price of Homes Sold", "Affordability Index - Mortgage", "Affordability Index - Rent")
new_varnames <- c("CSA2010", "mediansold12", "ai_mortgage12", "ai_rent12")

bnia2012 %<>%
  select(bnia_vars) %>%
  set_colnames(new_varnames)

#2013
bnia_vars <- c("CSA2010", "Median Price of Homes Sold", "Affordability Index - Mortgage", "Affordability Index - Rent")
new_varnames <- c("CSA2010", "mediansold13", "ai_mortgage13", "ai_rent13")

bnia2013 %<>%
  select(bnia_vars) %>%
  set_colnames(new_varnames)

#2014
bnia_vars <- c("CSA2010", "Median Price of Homes Sold", "Affordability Index - Mortgage", "Affordability Index - Rent")
new_varnames <- c("CSA2010", "mediansold14", "ai_mortgage14", "ai_rent14")

bnia2014 %<>%
  select(bnia_vars) %>%
  set_colnames(new_varnames)

#2015
bnia_vars <- c("CSA2010", "Median Price of Homes Sold", "Affordability Index - Mortgage", "Affordability Index - Rent")
new_varnames <- c("CSA2010", "mediansold15", "ai_mortgage15", "ai_rent15")

bnia2015 %<>%
  select(bnia_vars) %>%
  set_colnames(new_varnames)

#2016
bnia_vars <- colnames(bnia2016[, c("CSA2010", "Percent of Residents - White/Caucasian (Non-Hispanic)", "Percent of Residents - Hispanic", "Percent of Population 18-24 years old", "Median Price of Homes Sold", "Affordability Index - Mortgage", "Affordability Index - Rent")])
new_varnames <- c("CSA2010", "nonwhite16", "hispanic16", "percentya16", "mediansold16", "ai_mortgage16", "ai_rent16")

bnia2016 %<>%
  select(bnia_vars) %>%
  set_colnames(new_varnames) %>%
  mutate(nonwhite16 = (100 - nonwhite16))

#2017
bnia_vars <- colnames(bnia2017[, c("CSA2010", "Percent of Residents - White/Caucasian (Non-Hispanic)", "Percent of Residents - Hispanic", "Percent of Population 18-24 years old", "Median Price of Homes Sold", "Affordability Index - Mortgage", "Affordability Index - Rent")])
new_varnames <- c("CSA2010", "nonwhite17", "hispanic17", "percentya17", "mediansold17", "ai_mortgage17", "ai_rent17")

bnia2017 %<>%
  select(bnia_vars) %>%
  set_colnames(new_varnames) %>%
  mutate(nonwhite17 = (100 - nonwhite17))

```

```{r, merge BNIA to Zillow files}

RVBM %<>%
  left_join(bnia2010, by = "CSA2010") %>%
  left_join(bnia2011, by = "CSA2010") %>%
  left_join(bnia2012, by = "CSA2010") %>%
  left_join(bnia2013, by = "CSA2010") %>%
  left_join(bnia2014, by = "CSA2010") %>%
  left_join(bnia2015, by = "CSA2010") %>%
  left_join(bnia2016, by = "CSA2010") %>%
  left_join(bnia2017, by = "CSA2010")

HVBM %<>%
  left_join(bnia2010, by = "CSA2010") %>%
  left_join(bnia2011, by = "CSA2010") %>%
  left_join(bnia2012, by = "CSA2010") %>%
  left_join(bnia2013, by = "CSA2010") %>%
  left_join(bnia2014, by = "CSA2010") %>%
  left_join(bnia2015, by = "CSA2010") %>%
  left_join(bnia2016, by = "CSA2010") %>%
  left_join(bnia2017, by = "CSA2010")

```

```{r, prepare "long-format" demographic variables}


RVBM$nonwhite <- ifelse(RVBM$year == 2010, RVBM$nonwhite10,
                        ifelse(RVBM$year == 2016, RVBM$nonwhite16,
                               ifelse(RVBM$year == 2017, RVBM$nonwhite17, "NA")))

RVBM$hispanic <- ifelse(RVBM$year == 2010, RVBM$hispanic10,
                        ifelse(RVBM$year == 2016, RVBM$hispanic16,
                               ifelse(RVBM$year == 2017, RVBM$hispanic17, "NA")))

RVBM$percentya <- ifelse(RVBM$year == 2010, RVBM$percentya10,
                        ifelse(RVBM$year == 2016, RVBM$percentya16,
                               ifelse(RVBM$year == 2017, RVBM$percentya17, "NA")))

RVBM$mediansold <- ifelse(RVBM$year == 2010, RVBM$mediansold10,
                          ifelse(RVBM$year == 2011, RVBM$mediansold11,
                            ifelse(RVBM$year == 2012, RVBM$mediansold12,
                              ifelse(RVBM$year == 2013, RVBM$mediansold13,
                                ifelse(RVBM$year == 2014, RVBM$mediansold14,
                                  ifelse(RVBM$year == 2015, RVBM$mediansold15,
                                   ifelse(RVBM$year == 2016, RVBM$mediansold16,
                                      ifelse(RVBM$year == 2017, RVBM$mediansold17, "NA"))))))))

RVBM$ai_mortgage <- ifelse(RVBM$year == 2010, RVBM$ai_mortgage10,
                          ifelse(RVBM$year == 2011, RVBM$ai_mortgage11,
                            ifelse(RVBM$year == 2012, RVBM$ai_mortgage12,
                              ifelse(RVBM$year == 2013, RVBM$ai_mortgage13,
                                ifelse(RVBM$year == 2014, RVBM$ai_mortgage14,
                                  ifelse(RVBM$year == 2015, RVBM$ai_mortgage15,
                                   ifelse(RVBM$year == 2016, RVBM$ai_mortgage16,
                                      ifelse(RVBM$year == 2017, RVBM$ai_mortgage17, "NA"))))))))

RVBM$ai_rent <- ifelse(RVBM$year == 2010, RVBM$ai_rent10,
                          ifelse(RVBM$year == 2011, RVBM$ai_rent11,
                            ifelse(RVBM$year == 2012, RVBM$ai_rent12,
                              ifelse(RVBM$year == 2013, RVBM$ai_rent13,
                                ifelse(RVBM$year == 2014, RVBM$ai_rent14,
                                  ifelse(RVBM$year == 2015, RVBM$ai_rent15,
                                   ifelse(RVBM$year == 2016, RVBM$ai_rent16,
                                      ifelse(RVBM$year == 2017, RVBM$ai_rent17, "NA"))))))))

#HVBM
HVBM$nonwhite <- ifelse(HVBM$year == 2010, HVBM$nonwhite10,
                        ifelse(HVBM$year == 2016, HVBM$nonwhite16,
                               ifelse(HVBM$year == 2017, HVBM$nonwhite17, "NA")))

HVBM$hispanic <- ifelse(HVBM$year == 2010, HVBM$hispanic10,
                        ifelse(HVBM$year == 2016, HVBM$hispanic16,
                               ifelse(HVBM$year == 2017, HVBM$hispanic17, "NA")))

HVBM$percentya <- ifelse(HVBM$year == 2010, HVBM$percentya10,
                        ifelse(HVBM$year == 2016, HVBM$percentya16,
                               ifelse(HVBM$year == 2017, HVBM$percentya17, "NA")))

HVBM$mediansold <- ifelse(HVBM$year == 2010, HVBM$mediansold10,
                          ifelse(HVBM$year == 2011, HVBM$mediansold11,
                            ifelse(HVBM$year == 2012, HVBM$mediansold12,
                              ifelse(HVBM$year == 2013, HVBM$mediansold13,
                                ifelse(HVBM$year == 2014, HVBM$mediansold14,
                                  ifelse(HVBM$year == 2015, HVBM$mediansold15,
                                   ifelse(HVBM$year == 2016, HVBM$mediansold16,
                                      ifelse(HVBM$year == 2017, HVBM$mediansold17, "NA"))))))))

HVBM$ai_mortgage <- ifelse(HVBM$year == 2010, HVBM$ai_mortgage10,
                          ifelse(HVBM$year == 2011, HVBM$ai_mortgage11,
                            ifelse(HVBM$year == 2012, HVBM$ai_mortgage12,
                              ifelse(HVBM$year == 2013, HVBM$ai_mortgage13,
                                ifelse(HVBM$year == 2014, HVBM$ai_mortgage14,
                                  ifelse(HVBM$year == 2015, HVBM$ai_mortgage15,
                                   ifelse(HVBM$year == 2016, HVBM$ai_mortgage16,
                                      ifelse(HVBM$year == 2017, HVBM$ai_mortgage17, "NA"))))))))

HVBM$ai_rent <- ifelse(HVBM$year == 2010, HVBM$ai_rent10,
                          ifelse(HVBM$year == 2011, HVBM$ai_rent11,
                            ifelse(HVBM$year == 2012, HVBM$ai_rent12,
                              ifelse(HVBM$year == 2013, HVBM$ai_rent13,
                                ifelse(HVBM$year == 2014, HVBM$ai_rent14,
                                  ifelse(HVBM$year == 2015, HVBM$ai_rent15,
                                   ifelse(HVBM$year == 2016, HVBM$ai_rent16,
                                      ifelse(HVBM$year == 2017, HVBM$ai_rent17, "NA"))))))))

RVBM %<>% 
  mutate_at(c("nonwhite", "hispanic", "percentya", "mediansold", "ai_mortgage", "ai_rent"), as.numeric)

HVBM %<>% 
  mutate_at(c("nonwhite", "hispanic", "percentya", "mediansold", "ai_mortgage", "ai_rent"), as.numeric)

```

#Results

Maps comparing comparing Baltimore City distributions (across Community Statistical Areas) of different measures of housing values and affordability, as well as age and race distributions. Maps shown are for January 2010 and January 2017.

```{r}

bnia_sf %>%
  mutate(CSA2010 = Community) %>%
  left_join(HVBM, by = "CSA2010") %>%
  filter(year == c("2017", "2010")) %>%
  ggplot() +
    geom_sf(aes(fill = HomeValue)) +
    facet_wrap(~year)

```

```{r}
bnia_sf %>%
  mutate(CSA2010 = Community) %>%
  left_join(RVBM, by = "CSA2010") %>%
  filter(year == c("2017", "2010")) %>%
  ggplot() +
    geom_sf(aes(fill = RentalValue)) +
    facet_wrap(~year)

bnia_sf %>%
  mutate(CSA2010 = Community) %>%
  left_join(RVBM, by = "CSA2010") %>%
  filter(year == c("2017", "2010")) %>%
  ggplot() +
    geom_sf(aes(fill = mediansold)) +
    facet_wrap(~year)

bnia_sf %>%
  mutate(CSA2010 = Community) %>%
  left_join(RVBM, by = "CSA2010") %>%
  filter(year == c("2017", "2010")) %>%
  ggplot() +
    geom_sf(aes(fill = ai_mortgage)) +
    facet_wrap(~year)

bnia_sf %>%
  mutate(CSA2010 = Community) %>%
  left_join(RVBM, by = "CSA2010") %>%
  filter(year == c("2017", "2010")) %>%
  ggplot() +
    geom_sf(aes(fill = ai_rent)) +
    facet_wrap(~year)

bnia_sf %>%
  mutate(CSA2010 = Community) %>%
  left_join(RVBM, by = "CSA2010") %>%
  filter(year == c("2017", "2010")) %>%
  ggplot() +
    geom_sf(aes(fill = percentya)) +
    facet_wrap(~year)

bnia_sf %>%
  mutate(CSA2010 = Community) %>%
  left_join(RVBM, by = "CSA2010") %>%
  filter(year == c("2017", "2010")) %>%
  ggplot() +
    geom_sf(aes(fill = nonwhite)) +
    facet_wrap(~year)


```

Visualizations for H1: County-level rental values will show a greater increase over time than housing values over the same period

```{r}

year_breaks <- c("2010-01", "2011-01", "2012-01", "2013-01", "2014-01", "2015-01", "2016-01", "2017-01", "2018-01")

combinedMD %>%
  mutate(hrv_ratio = (RentalValue/(HomeValue/1000))) %>%
  filter(Metro.x %in% "Baltimore-Columbia-Towson") %>%
  ggplot() +
  geom_jitter(aes(x = date, y = hrv_ratio , group = RegionName, color = RegionName)) + 
  ggtitle("Change in Rental-to-Homevalue Ratio Over Time in Baltimore Metro") + 
  scale_x_discrete("Date", c('2010', year_breaks, year_breaks)) + 
  ylab("Ratio of rental values to home values")

combinedMD %>%
  filter(Metro.x %in% "Baltimore-Columbia-Towson") %>%
  ggplot() +
  geom_jitter(aes(x = date, y = RentalValue , group = RegionName, color = RegionName)) + 
  ggtitle("Rental Values Over Time in Baltimore Metro") + 
  scale_x_discrete("Date", c('2010', year_breaks, year_breaks)) + 
  ylab("Average rental Values (monthly rent in $)")

combinedMD %>%
  filter(Metro.x %in% "Baltimore-Columbia-Towson") %>%
  ggplot() +
  geom_jitter(aes(x = date, y = (HomeValue/1000) , group = RegionName, color = RegionName)) + 
  ggtitle("Home Values Ratio Over Time in Baltimore Metro") + 
  scale_x_discrete("Date", c('2010', year_breaks, year_breaks)) + 
  ylab("Average home value (in thousands)")

```

Visualizations for H2: This effect will be more pronounced in urban than in rural counties

```{r}

#map filling MD counties by average rental values in 2017
HVMD %>%
    left_join(RVMD, by = c("RegionName", "date")) %>%
    mutate(hrv_ratio = (RentalValue/(HomeValue/1000))) %>%
    mutate(subregion = gsub(" County", "", RegionName)) %>%
    mutate(subregion = tolower(subregion)) %>%
    right_join(md_map, by = "subregion") %>%
    filter(date %in% c("2018-12", "2015-12")) %>%
    ggplot() + 
      geom_polygon(aes(x = long, y = lat, group = subregion, fill = hrv_ratio)) +
      facet_wrap(~date, scale = "free") +
      ggtitle("MD Rental-to-HomeValue Ratios in Dec. 2015 and Dec. 2018") +
      scale_fill_continuous(name = "RTHV Ratio")

#map filling MD counties by average rental values in 2017
HVMD %>%
    left_join(RVMD, by = c("RegionName", "date")) %>%
    mutate(hrv_ratio = (RentalValue/(HomeValue/1000))) %>%
    mutate(subregion = gsub(" County", "", RegionName)) %>%
    mutate(subregion = tolower(subregion)) %>%
    right_join(md_map, by = "subregion") %>%
    filter(date %in% c("2018-12", "2015-12")) %>%
    ggplot() + 
      geom_polygon(aes(x = long, y = lat, group = subregion, fill = hrv_ratio)) +
      facet_wrap(~date, scale = "free") +
      ggtitle("MD Rental-to-HomeValue Ratios in Dec. 2015 and Dec. 2018") +
      scale_fill_continuous(name = "RTHV Ratio")
```

Visualizations for H3: Areas with younger populations will have higher rental values (compared to housing values) than areas with older populations

```{r}
#Visualize relationships between age and race make-up of community and average home and rent values in community. Charts created with both Zillow data and BNIA data for comparison purposes. Data for 2016-17

HVBM %>%
  filter(year %in% c("2016", "2017")) %>%
  mutate(nw_binary = ifelse(nonwhite <= 50, 0, 1)) %>%
  ggplot() +
  geom_jitter(aes(x = percentya, y = HomeValue/1000 , group = RegionName, color = as.factor(nw_binary))) + ggtitle("Zillow Home Value by Percent of Population 18-24 Years Old") + xlab("Percent of Population 18-24 Years Old") + ylab("Average home value (in thousands)") +
  scale_color_discrete(name = "Percentage of population nonwhite", labels = c("More than 50% nonwhite", "50% nonwhite or less"))


RVBM %>%
  filter(year %in% c("2016", "2017")) %>%
  mutate(nw_binary = ifelse(nonwhite <= 50, 0, 1)) %>%
  ggplot() +
  geom_jitter(aes(x = percentya, y = RentalValue , group = RegionName, color = as.factor(nw_binary))) +
  ggtitle("Zillow Rental Value by Percent of Population 18-24 Years Old") + 
  xlab("Percent of Population 18-24 Years Old") + 
  ylab("Zillow Rental Value") +
  scale_color_discrete(name = "Percentage of population nonwhite", labels = c("More than 50% nonwhite", "50% nonwhite or less"))

RVBM %>%
  filter(year %in% c("2016", "2017")) %>%
  mutate(nw_binary = ifelse(nonwhite <= 50, 0, 1)) %>%
  ggplot() +
  geom_jitter(aes(x = percentya, y = ai_mortgage , group = RegionName, color = as.factor(nw_binary))) +
  ggtitle("Mortgage affordability index by Percent of Population 18-24 Years Old") +
  xlab("Percent of Population 18-24 Years Old") + 
  ylab("Mortgage Affordability Index") +
  scale_color_discrete(name = "Percentage of population nonwhite", labels = c("More than 50% nonwhite", "50% nonwhite or less"))

RVBM %>%
  filter(year %in% c("2016", "2017")) %>%
  mutate(nw_binary = ifelse(nonwhite <= 50, 0, 1)) %>%
  ggplot() +
  geom_jitter(aes(x = percentya, y = ai_rent, group = RegionName, color = as.factor(nw_binary))) + 
  ggtitle("Rent Affordability Index by Percent of Population 18-24 Years Old") + xlab("Percent of Population 18-24 Years Old") +
  ylab("Rent Affordability Index") +
  scale_color_discrete(name = "Percentage of population nonwhite", labels = c("More than 50% nonwhite", "50% nonwhite or less"))

```

Visualizations for H4: Increases over time in home and rental values will be associated with decreases in the proportion of non-white residents

```{r}

```

#Discussion

While Zillow's housing data proved useful for mapping at the state/county level, and showed similar distributions across communities to the BNIA data, there were a number of disadvantages to working with this data. There were many missing values that could interrupt trend lines and produced blank areas when mapping -- in particular this is visible in the map of HousingValue during 2010, where the lower-income, less-white portions of the city are largely missing. This suggests that the potential for bias in the Zillow data may increase the further back the data goes -- this makes sense, as one would expect their database to become more robust over time, as use of the site has increased and techniques for data mining have improved.

Another major issue we encountered with the Zillow datasets is the lack of standardized geographic identifiers for merging with other data and mapping at meaningful community levels -- Zillow data had to be merged to both ACS and BNIA data by county or neighborhood name, and these text-based identifiers proved to be significantly more labor-intensive to preprocess. Zillow's neighborhood names also had significant alignment issues with the only publicly-available Baltimore neighborhood shapefile, leading to large blank areas when mapping was attempted.

Neighborhood names in general proved to be problematic to work with -- in fact, we could not obtain demogrpahic data for areas smaller than the Community Statistical Areas, which we were able to map to neighborhoods using BNIA's provided geographic crosswalk, but encountered many problems in merging due to misspellings, neighborhoods with multiple names, neighborhoods that are sometimes counted as part of a larger surrounding neighborhood, and abbreviations.

Finally, we found it difficult to locate comparable measures to Zillow's home value and median rental value indexes, making it difficult to assess the validity of these measures. We do have some evidence of their validity at least in terms of comparing different geographic areas, as the Zillow data displayed similar trends to the BNIA data when mapped onto Baltimore city by neighborhood/CSA. However, it would not be possible to make a general statement about the validity of these measures based 

This exploration of exploring housing and demographic relationships effectively illustrates the difficulties, limitations, and extra processing burden of working with open access data. However, the Zillow data did allow us to compare market estimates to survey-based estimates, visualize seasonal changes in housing & rental markets, and provided a more direct estimate of rental markets than was available from any of our other data sources.


#References

Brummet, Q. and Reed D., (2019), The Effects of Gentrification on the Well-Being and Opportunity of Original Resident Adults and Children. FRB of Philadelphia Working Paper No. 19-30.

Capps, K., (2019, July 16), The Hidden Winners in Neighborhood Gentrification. Retrieved from https://www.citylab.com/equity/2019/07/gentrification-effects-neighborhood-data-economic-statistics/594064/.

Hoffower, H., (2019, July 29), Millennials aren't buying homes, and it might not be because they can't afford them: Some actually prefer to rent instead. Retrieved from https://www.businessinsider.com/millennials-renting-homes-instead-of-buying-2019-7.

Hoynes, H.W. and McFadden, D. (1994), The Impact of Demographics on Housing and Non-Housing Wealth in the United States. NBER Working Paper No. 4666.

Kasakove, S., (2019, September 25), AOC’s Plan to Decommodify Housing. Retrieved from https://www.thenation.com/article/aoc-sanders-housing-rent/.

Meehan, S., (2019, March 20), Baltimore among nation’s most gentrified cities, study shows. Retrieved from https://www.baltimoresun.com/maryland/baltimore-city/bs-md-ci-gentrification-study-20190319-story.html.

Squires, G. and Kubrin, C., (2005), Privileged Places: Race, Uneven Development, and the Geography of Opportunity in Urban America. Urban Studies, 42.


